---
title: "Lab 5"
author: "Lamae Maharaj"
output: pdf_document
---


We will work with the diamonds dataset from last lecture:

```{r}
pacman::p_load(ggplot2) #this loads the diamonds data set too
?diamonds
diamonds$cut =      factor(diamonds$cut, ordered = FALSE)
diamonds$color =    factor(diamonds$color, ordered = FALSE)
diamonds$clarity =  factor(diamonds$clarity, ordered = FALSE)
skimr::skim(diamonds)
```

Given the information above, what are the number of columns in the raw X matrix?

There are 10 columns in the raw X matrix. This can be confirmed by using ncol(MATRIX) or by viewing the summary above.

Verify this using code:

```{r}
ncol(diamonds)
```

Would it make sense to use polynomial expansions for the variables cut, color and clarity? Why or why not?

No it would not makes sense to use polynomial expansions for these variables since they are categorical (nominal and ordinal). It does not matter of whether or not they are nominal or ordinal. Due to the fact that they are categorical can make it so that a polynomial expansion will have large amounts of error.


Would it make sense to use log transformations for the variables cut, color and clarity? Why or why not?

No because there is no way to take a log of a non numerical value. 

In order to ensure there is no time trend in the data, randomize the order of the diamond observations in D:.

```{r}
length = dim(diamonds)[1]
randObservations = sample(nrow(diamonds), length)
diamonds = diamonds[randObservations,]
ggplot(diamonds, aes(x = carat, y = price)) + geom_point()
```

Let's also concentrate only on diamonds with <= 2 carats to avoid the issue we saw with the maximum. So subset the dataset. Create a variable n equal to the number of remaining rows as this will be useful for later. Then plot it.

```{r}
#TO-DO
caratSubset = diamonds[which(diamonds$carat<=2),]
n = nrow(diamonds) - nrow(caratSubset)
ggplot(caratSubset, aes(x = carat, y = price)) +  geom_point()
```

Create a linear model of price ~ carat and gauge its in-sample performance using s_e.

```{r}
x = caratSubset$carat
y = caratSubset$price
pricePerCarat = lm(y~x)
standardError = sd(pricePerCarat$residuals)
standardError
```

Create a model of price ~ clarity and gauge its in-sample performance

```{r}
x = caratSubset$clarity
y = caratSubset$price
priceClarity = lm(y~x)
se_0 = sd(priceClarity$residuals)
se_0
```

Why is the model price ~ carat substantially more accurate than price ~ clarity?

The model price ~ carat is substantially more accurate because its simply a more "important" feature in our phenomena. 
Clarity can also be a categorical variable, this means that fitting a line to this will most likely have more error than fitting to a continuous variable. 

Create a new transformed feature ln_carat and plot it vs price.

```{r}
#TO-DO
caratSubset = transform(caratSubset, ln_carat = log(carat))
ln_carat = log(caratSubset$carat)
ggplot(caratSubset, aes(x = ln_carat, y = price)) + 
  geom_point()
```

Would price ~ ln_carat be a better fitting model than price ~ carat? Why or why not?

It depends what sort of model that we are fitting. If it is a linear one then it would not make sense to use the log transformation. If there is a use of a second order polynomial, then the log transformation would be a better option.

Verify this by comparing R^2 and RMSE of the two models:

```{r}
mod1 = lm(caratSubset$price ~ caratSubset$carat)
mod2 = lm(caratSubset$price ~ caratSubset$ln_carat)
mod1RMSE = sqrt(mean(mod1$residuals^2))
mod1R_SQ = summary(mod1)$r.squared
mod2RMSE = sqrt(mean(mod2$residuals^2))
mod2R_SQ = summary(mod2)$r.squared
mod1RMSE
mod1R_SQ
mod2RMSE
mod2R_SQ
```

Create a new transformed feature ln_price and plot its estimated density:


```{r}
#TO-DO
caratSubset = transform(caratSubset, ln_price = log(price))
ggplot(caratSubset) + geom_histogram(aes(x = ln_price), binwidth = 0.01)
```


Now plot it vs carat.

```{r}
ggplot(diamonds, aes(x = carat, y = ln_price)) +  geom_point()
```

Would ln_price ~ carat be a better fitting model than price ~ carat? Why or why not?

No it would not be a better fit model. Due to this being a relationship that is going to have larger and larger residuals the farther out in carat we can predict for.

Verify this by computing s_e of this new model. Make sure these metrics can be compared apples-to-apples with the previous.

```{r}
logPricePerCarat = lm(caratSubset$ln_price ~ caratSubset$carat)
logRMSE = (sqrt(mean(logPricePerCarat$residuals^2))) 
logYhat = logPricePerCarat$fitted.values
yHat = exp(logYhat)
e = caratSubset$price - yHat
sse = sum(e^2)
sst = sum((caratSubset$price - mean(caratSubset$price))^2)
logPricePerCaratRMSE = sqrt(sse/ (nrow(caratSubset)-2))
```

We just compared in-sample statistics to draw a conclusion on which model has better performance. But in-sample statistics can lie! Why is what we did valid?

What we did is valid because we are evaluating the fit of the line without over-fitting. There is a comparison of two models that are using the same feature against the price. Transforming error so that they are both in the same units. 

Plot ln_price vs ln_carat.

```{r}
ggplot(diamonds, aes(x = ln_carat, y = ln_price)) + geom_point()
```

Would ln_price ~ ln_carat be the best fitting model than the previous three we considered? Why or why not?

Verify this by computing s_e of this new model. Make sure these metrics can be compared apples-to-apples with the previous.

```{r}
logModel = lm(caratSubset$ln_price ~ caratSubset$ln_carat)
```

Compute b, the OLS slope coefficients for this new model of ln_price ~ ln_carat.

```{r}
logModel$coefficients
```

Interpret b_1, the estimated slope of ln_carat.

Every 1 unit(log units) change for carat their is ~69% change in the price of a diamond

Interpret b_0, the estimated intercept.

With a 0 carat diamond in log units

Create other features ln_x, ln_y, ln_z, ln_depth, ln_table.

```{r}
caratSubset = transform(caratSubset, ln_x = log(caratSubset$x))
caratSubset = transform(caratSubset, ln_y = log(caratSubset$y))
caratSubset = transform(caratSubset, ln_z = log(caratSubset$z))
caratSubset = transform(caratSubset, ln_depth = log(caratSubset$depth))
caratSubset = transform(caratSubset, ln_table = log(caratSubset$table))
```

From now on, we will be modeling ln_price (not raw price) as the prediction target. 

Create a model (B) of ln_price on ln_carat interacted with clarity and compare its performance with the model (A) ln_price ~ ln_carat.

```{r}
ggplot(caratSubset, aes(x = ln_carat, y = ln_price, color = clarity)) + geom_point()
moda = lm(ln_price ~ ln_carat, caratSubset)
modb = lm(ln_price ~ ln_carat * clarity, caratSubset)
summary(moda)$sigma
summary(modb)$sigma
```

Which model does better? Why?

Model B does better. Due to the fact that this is the result of interacting the feature ln_carat with clarity and other variables . 

Create a model of (C) ln_price on ln_carat interacted with every categorical feature (clarity, cut and color) and compare its performance with model (B)

```{r}
#Model C
modc = lm(ln_price ~ (ln_carat) * (clarity + cut  + color), caratSubset)
summary(modc)$sigma
summary(modb)$sigma
```

Which model does better? Why?

 Model C does better since it is interacting between features clarity, cut and color.

Create a model (D) of ln_price on every continuous feature (logs of carat, x, y, z, depth, table) interacted with every categorical feature (clarity, cut and color) and compare its performance with model (C).

```{r}
#Model D
caratSubset2  = caratSubset[which(caratSubset$x>0 & caratSubset$y>0 & caratSubset$z>0),]
modd = lm(ln_price ~ (ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table) * (clarity + cut  + color), caratSubset2)
summary(modd)$sigma
summary(modc)$sigma
```

Which model does better? Why?

Model D does bettor because now we are capturing the interactions between 6 features  by  3 features and 
1 feature by 3 features as shown in Model C. 

What is the p of this model D? Compute with code.

```{r}
modd$rank
ncol(model.matrix(~ (ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table) * (clarity + cut  + color), caratSubset2))
```

Create model (E) which is the same as before except create include the raw features interacted with the categorical features and gauge the performance against (D).

```{r}
#Model E
mod_e = lm(ln_price ~ (carat + x + y + z + depth + table) * (clarity + cut  + color), caratSubset2)
summary(mod_e)$sigma
summary(modd)$sigma
```

Which model does better? Why?

Model E performs better. However, the returns are diminishing. 
There are no longer uses of the log transformations, but now are use the raw features are in effect. 

Create model (F) which is the same as before except also include also third degree polynomials of the continuous features interacted with the categorical features and gauge performance against (E). By this time you're getting good with R's formula syntax!

```{r}
#Model F
modf = lm(ln_price ~ (poly(carat,3) + poly(x,3) + poly(y,3) + poly(z,3) + poly(depth,3) + poly(table,3)) * (clarity + cut  + color), caratSubset2)
summary(modf)$sigma
summary(mod_e)$sigma
```

Which model does better? Why?

Model F does better, but the cost of its performance is unknown. This can be forever continuing. 

Can you think of any other way to expand the candidate set curlyH? Discuss.

An expansion of curlyH would be to increase the order of the polynomial.
Another way would be to use different transformations that are not linearly dependent on each other.

We should probably assess oos performance now. Sample 2,000 diamonds and use these to create a training set of 1,800 random diamonds and a test set of 200 random diamonds. Define K and do this splitting:

```{r}
K = 10
set.seed(1984)
sampleNum = 2000
splitIndex = nrow(D)*(1-1/K)
D = caratSubset2[sample(1:nrow(caratSubset2), sampleNum), ]
Dtrain = D[1:splitIndex,]
Dtest = D[(splitIndex+1):nrow(D),]
```

Compute in and out of sample performance for models A-F. Use s_e as the metric (standard error of the residuals). Create a list with keys A, B, ..., F to store these metrics. Remember the performances here will be worse than before since before you're using nearly 52,000 diamonds to build a model and now it's only 1,800! 

```{r}
inSampleSE = list()
outSampleSE = list()
moda = lm(ln_price ~ ln_carat, Dtrain)
inSampleSE[['A']] = sd(moda$residuals)
inSampleSE[['B']] = sd(modb$residuals)
inSampleSE[['C']] = sd(modc$residuals)
inSampleSE[['D']] = sd(modd$residuals)
inSampleSE[['E']] = sd(mod_e$residuals)
inSampleSE[['F']] = sd(modf$residuals)
outSampleSE[['A']] = sd(Dtest$ln_price - predict(moda,Dtest))
outSampleSE[['B']] = sd(Dtest$ln_price - predict(modb,Dtest))
outSampleSE[['C']] = sd(Dtest$ln_price - predict(modc,Dtest))
outSampleSE[['D']] = sd(Dtest$ln_price - predict(modd,Dtest))
outSampleSE[['E']] = sd(Dtest$ln_price - predict(mod_e,Dtest))
outSampleSE[['F']] = sd(Dtest$ln_price - predict(modf,Dtest))
sd(priceClarity$residuals)
unlist(inSampleSE)
unlist(outSampleSE)
```

You computed oos metrics only on n_* = 200 diamonds. What problem(s) do you expect in these oos metrics?

These metrics are not accurate becuase only the testing data is being used. 

To do the K-fold cross validation we need to get the splits right and crossing is hard. I've developed code for this already. Run this code.

```{r}
temp = rnorm(n)
folds_vec = cut(temp, breaks = quantile(temp, seq(0, 1, length.out = K + 1)), include.lowest = TRUE, labels = FALSE)
head(folds_vec, 200)
```

Comment on what it does and how to use it to do a K-fold CV:

#TO-DO

Do the K-fold cross validation for model F and compute the overall s_e and s_s_e. 

```{r}
oos_SE = array(NA, K) #Setup an initial array for errors
yHat = array(NA, nrow(caratSubset2))
for (k in 1 : K){
  testIndex = which(folds_vec == k)
  trainIndex = setdiff(1 : nrow(caratSubset2), testIndex)
  
  modf = lm(ln_price ~ (poly(carat,3) + poly(x,3) + poly(y,3) + poly(z,3) + poly(depth,3) + poly(table,3)) * (clarity + cut  + color), caratSubset2[trainIndex,])
  yHat[testIndex] = predict(modf, caratSubset2[testIndex,])
  
  oos_SE[k] = sd(caratSubset2[testIndex,]$ln_price - yHat[testIndex]) 
}
oos_SE
sd(oos_SE)
mean(oos_SE)
```

Does K-fold CV help reduce variance in the oos s_e? Discuss.

K-Fold helps reduce the variance in the dataset partitions which then gives a truthful representation of the oos metrics.

Imagine using the entire rest of the dataset besides the 2,000 training observations divvied up into slices of 200. Measure the oos error for each slice on Model F in a vector s_e_s_F and compute the s_s_e_F and also plot it.

```{r}
#TO-DO
ggplot(data.frame(s_e_s_F = s_e_s_F)) + geom_histogram(aes(x = s_e_s_F))
```

